The Regional Effects of Marine Protected Areas

### 2020-03-12 documents/ovando-regional-effects-of-mpas-v2.Rmd
passing to abalone

@@

@@

channel_islands <- readRDS(here::here("data","channel_islands_map.RDS"))

ca_mpas <- sf::st_read(here::here("data","MPA_CA_Existing_160301")) %>%
  rmapshaper::ms_simplify() %>%
  sf::st_transform(crs = 4326)

@@

As stakeholders around the world increasingly seek to use MPAs in marine resource management portfolios, and the performance of existing MPAs is evaluated, it is critical that we develop a better understanding of the magnitude and drivers of regional-scale MPA effects. The Channel Islands National Marine Sanctuary, California, USA provides an ideal case study to address this need. A network of protected areas covering approximately 20% of the Islands' waters qas put in place in 2003 (see @osmond2010, @kirlin2013, @botsford2014, and @hilborn2012 for information on the creation of these MPAs). We use data from the first 15 years of protection to provide what is to our knowledge the first large-scale empirical assessment of the regional effect of a large MPA network on a wide array of fin-fish species. In contrast to clear differences in biomass densities observed inside and outside of well protected MPAs [@lester2009], we are unable to detect a clear regional effect from the Channel Islands MPAs. We build off of existing MPA theory to consider why this might be, and provide guidance for scientists and managers as to when and how we might expect to estimate the regional conservation effects of MPAs.

@@

Numerous factors can affect the regional effects of MPAs. These include the scale of adult and larval dispersal relative to the size of the MPAs  [@gaines2003; @botsford2008; @difranco2018;@mcgilliard2008], larval dispersal patternss and the strength and timing of density dependence in the population (e.g. pre- or post-settlement, @burgess), how overfished the population would be without the MPA, and how fishing activity and management responds to the implementation of the MPAs [@hilborn2004; @hilborn1992; @walters2004; @hastings2003;@gerber2003; @gerber2005; @hilborn2004a; @gaines2010; @botsford2003;@smith2003;@smith2006]. In addition, even for the same total area of MPAs, the location and spacing of the MPAs can have a profound influence on their cumulative impact through habitat and network effects [@gaines2010; @costello2010].

@@

Although the Islands do not encompass the full population of any of these species, they encompass a large enough area to assess the regional MPA conservation effects we might see in the Channel Islands after 13 years of protection (the time span covered by our model).  The median regional conservation effect produced by our simulation model after 13 years of protection by MPAs covering between 15%-20% of a population's range was `r percent(med_ci_effect, accuracy = 1)`. This projected effect has an an interquartile range of `r percent(quart_ci_effect[2], accuracy = 1)` and `r percent(quart_ci_effect[4], accuracy = 1)`, although values as high as a 200% increase and as low as a 50% decrease were also produced. Our expectation then is that all else being equal we are likely looking for a modest but positive effect size.

@@

What evidence do we have  for the effects of the Channel Island MPAs? @caslle2015 examined changes in biomass densities of species targeted by fishing activity inside and outside of the MPAs over time (response ratios), and compared these changes to comparable trends in non-targeted species. They found a statistically significant increase in the response ratios of targeted speices over time, and *evidenceXX* that this increase is smaller in the non-targeted species. Updating the results of @caselle2015 through 2018 shows a continuation of this trend (Fig.\@ref(fig:XX)).

@@

@@

ci_map <- pisco_data %>%
  left_join(site_data, by = c("site","side")) %>%
  filter(is.na(eventual_mpa) == F) %>%
  filter(region %in% c("ANA", "SCI","SRI",'SMI'),
         classcode %in% top_species) %>%
  group_by(lat_wgs84, lon_wgs84, site, region, eventual_mpa) %>%
  count()

ci_map <-  ci_map %>%
  dplyr::mutate(geometry = purrr::map2(lon_wgs84, lat_wgs84, ~ sf::st_point(x = c(.x, .y), dim = 'XY'))) %>%
  ungroup() %>%
  mutate(geometry = sf::st_sfc(geometry, crs = 4326)) %>%
  sf::st_sf()

bbox <- sf::st_bbox(ci_map)

pisco_ci_map_plot <-  ggmap(channel_islands, alpha = 0.5) +
  geom_sf(data = ca_mpas, inherit.aes = FALSE, alpha = 0.5) +
  geom_sf(data = ci_map,inherit.aes = FALSE,
  aes(fill = eventual_mpa, color = eventual_mpa),
  shape = 21, 
  alpha = 0.5,
  show.legend = FALSE,
  size = 1.5) +
  coord_sf(xlim = c(bbox['xmin'] - .3, bbox['xmax'] + .06),
  ylim = c(bbox['ymin'] - .1, bbox['ymax'] + .1)) +
  # scale_color_gradient(low = "white", high = "orangered",
  # guide = guide_colorbar(frame.colour = "black",frame.linewidth = 1,
  # barheight = unit(13, "lines")),
  # name = "Observations")  +
  labs(x = "", y = "") +
      ggspatial::annotation_scale(location = "br", width_hint = 0.4, text_col = "white") + 
  ggsci::scale_color_npg() + 
    ggsci::scale_color_npg()

>>>>

  load(file.path(run_dir, 'abundance_data.Rdata'))

@@

load(file = file.path(run_dir,"plots.RData"))

@@

@@

As stakeholders around the world increasingly seek to use MPAs in marine resource management portfolios, and the performance of existing MPAs is evaluated, it is critical that we develop a better understanding of the magnitude and drivers of regional-scale MPA effects. The Channel Islands National Marine Sanctuary, California, USA provides an ideal case study to address this need. A network of protected areas covering approximately 20% of the Islands' waters qas put in place in 2003 (see @osmond2010, @kirlin2013, @botsford2014, and @hilborn2012 for information on the creation of these MPAs). We use data from the first 14 years of protection to provide what is to our knowledge the first large-scale empirical assessment of the regional effect of a large MPA network on a wide array of fin-fish species. In contrast to clear differences in biomass densities observed inside and outside of well protected MPAs [@lester2009], we are unable to detect a clear regional effect from the Channel Islands MPAs. We build off of existing MPA theory to consider why this might be, and provide guidance for scientists and managers as to when and how we might expect to estimate the regional conservation effects of MPAs.

@@

Numerous factors can affect the regional effects of MPAs. These include the scale of adult and larval dispersal relative to the size of the MPAs  [@gaines2003; @botsford2008; @difranco2018;@mcgilliard2008], larval dispersal patternss and the strength and timing of density dependence in the population (e.g. pre- or post-settlement, @burgess), how overfished the population would be without the MPA, and how fishing and management responds to the implementation of the MPAs [@hilborn2004; @hilborn1992; @walters2004; @hastings2003;@gerber2003; @gerber2005; @hilborn2004a; @gaines2010; @botsford2003;@smith2003;@smith2006]. In addition, even for the same total area of MPAs, the location and spacing of the MPAs can have a profound influence on their cumulative impact through habitat and network effects [@gaines2010; @costello2010].

@@

Although the Islands do not encompass the full population of any of these species, they encompass a large enough area to assess the regional MPA conservation effects we might see in the Channel Islands after 14 years of protection (the time span covered by our model).  The median regional conservation effect produced by our simulation model after 13 years of protection by MPAs covering between 15%-20% of a population's range was `r percent(med_ci_effect, accuracy = 1)`. This projected effect has an an interquartile range of `r percent(quart_ci_effect[2], accuracy = 1)` and `r percent(quart_ci_effect[4], accuracy = 1)`, although values as high as a 200% increase and as low as a 50% decrease were also produced. Our expectation then is that all else being equal we are likely looking for a modest but positive effect size.

@@

What evidence do we have  for the effects of the Channel Island MPAs? @caslle2015 examined changes in biomass densities of species targeted by fishing activity inside and outside of the MPAs over time (generally termed "response ratios"), and compared these changes to the trends in non-targeted species. They found a statistically significant increase in the response ratios of targeted speices over time, and evidence that this increase is smaller in the non-targeted species. Updating the results of @caselle2015 through 2018 shows a continuation of this trend (Fig.\@ref(fig:XX)).

@@

density_ratios %>% 
  ggplot(aes(years_protected, true_density_ratio)) + 
  geom_point()

@@

sample_sites <- pisco_data %>%
  left_join(site_data, by = c("site","side")) %>%
  filter(is.na(eventual_mpa) == F) %>%
  filter(region %in% c("ANA", "SCI","SRI",'SMI'),
         classcode %in% top_species) %>%
  group_by(lat_wgs84, lon_wgs84, site, region, eventual_mpa) %>%
  count()

channel_islands <- sf::st_read(here::here("data","cinms_py2","cinms_py.shp"))

rnaturalearth::ne_coastline()

ci_bbox <- sf::st_bbox(channel_islands)

sample_sites <-  sample_sites %>%
  st_as_sf(coords = c("lon_wgs84", "lat_wgs84"),
           crs = 4326)

california <- rnaturalearth::ne_states(country = "united states of america", returnclass = "sf") %>% 
  filter(name == "California") %>% 
    sf::st_transform(crs = sf::st_crs(channel_islands))

channel_islands_mpas <- sf::st_read(here::here("data","MPA_CA_Existing_160301"))%>%
  sf::st_transform(crs = sf::st_crs(channel_islands))

pisco_ci_map_plot <-  ggplot() + 
  geom_sf(data = california, fill = "lightgrey") +
  geom_sf(data = channel_islands_mpas, fill = "lightblue") +
 geom_sf(data = sample_sites,shape = 21, fill = "darkgrey",
                                      size = 2, alpha = .75) +

  # scale_color_gradient(low = "white", high = "orangered",
  # guide = guide_colorbar(frame.colour = "black",frame.linewidth = 1,
  # barheight = unit(13, "lines")),
  # name = "Observations")  +
  labs(x = "", y = "") +
  scale_y_continuous(limits = c(33.8, 34.2)) +
  scale_x_continuous(limits = c(-120.75,-119.25)) +
      ggspatial::annotation_scale(location = "br") + 
  #   coord_sf(xlim = c(bbox['xmin'] - .3, bbox['xmax'] + .06),
  # ylim = c(bbox['ymin'] - .1, bbox['ymax'] + .1)) +
    ggsci::scale_color_npg()
